Descriptive Complexity for Counting Complexity Classes

### 2017-07-24 LMCS/counting.tex
Paso completo a sección 4.

At this point it is natural to ask whether one can extend the previous idea to capture $\fpspace$~\cite{Ladner89}, the class of functions computable in polynomial space. 
Of course, for capturing this class one needs a logical core powerful enough, like $\pfp$, for simulating the run of a polynomial-space TM.
Moreover, 
one also needs more powerful quantitative quantifiers as functions like $2^{2^n}$ can be computed in polynomial space,
so second-order sum is not enough for the quantitative layer of a logic for $\fpspace$.
In fact, by considering second-order product we obtain the fragment $\qso(\pfp)$ that captures $\fpspace$. 
\begin{thm} \label{theo:capture-fpspace}
	$\qso(\pfp)$ captures $\fpspace$ over ordered structures.
\end{thm}
\proof
\input{proof-capture-fpspace}
\qed
%The proof of the previous theorem follows the same line as for the logical characterization of $\fp$: one shows that each function in $\qso(\pfp)$ can be computed in $\fpspace$ and, conversely, the output of a polynomial-space TM can be reconstructed by using $\pfp$ and quantitative quantifiers.

>>>>

At this point it is natural to ask whether one can extend the previous idea to capture $\fpspace$~\cite{Ladner89}, the class of functions computable in polynomial space. 
Of course, for capturing this class one needs a logical core powerful enough, like $\pfp$, for simulating the run of a polynomial-space TM.
Moreover, 
one also needs more powerful quantitative quantifiers as functions like $2^{2^n}$ can be computed in polynomial space,
so $\eqso$ is not enough for the quantitative layer of a logic for $\fpspace$.
In fact, by considering second-order product we obtain the fragment $\qso(\pfp)$ that captures $\fpspace$. 
\begin{thm} \label{theo:capture-fpspace}
	$\qso(\pfp)$ captures $\fpspace$ over ordered structures.
\end{thm}
\proof
\input{proof-capture-fpspace}
\qed
%The proof of the previous theorem follows the same line as for the logical characterization of $\fp$: one shows that each function in $\qso(\pfp)$ can be computed in $\fpspace$ and, conversely, the output of a polynomial-space TM can be reconstructed by using $\pfp$ and quantitative quantifiers.

### 2017-07-24 LMCS/extentions.tex
Paso completo a sección 4.

There exist complexity classes that do not fit in our framework because either the output of a function in not a natural number (e.g. a negative number) or the class is not defined purely in terms of arithmetical operations (e.g. min and max are used).
To remedy this problem, we show here how $\qso$ can be easily extended  to capture such classes that go beyond sum and product over natural numbers.

It is well-known that, under some reasonable complexity-theoretical assumptions, $\shp$ is not closed under subtraction, not even under subtraction by one~\cite{OH93}.
To overcome this limitation, $\gp$ was introduced in~\cite{FFK94} as the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x) - \tmr_M(x)$, where  $\tmr_M(x)$ is the number of rejecting runs of $M$ with input $x$.
That is, $\gp$ is the closure of $\shp$ functions under subtraction, and its functions can obviously take negative values.
Given that our logical framework was built on top of the natural numbers, we need to extend $\qso$ in order to capture $\gp$. 
The most elegant way to do this is by allowing constants coming from $\bbZ$ instead of just $\bbN$. 
Formally, we define the logic $\qsoz$ whose syntax is the same as in \eqref{syntax} and whose semantics is the same as in Table~\ref{tab-semantics} except that the atomic formula $s$ (i.e. a constant) comes from $\bbZ$.  
Similar than for $\qso$, we define the fragment $\eqsoz$ as the extension of $\eqso$ with constants in $\bbZ$.
\begin{exa}
	Recall the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the number of cliques in a graph that are not triangles. This can be easily done in $\qsoz$ with the formula:
	$
	\alpha_5 :=	\alpha_2 + (-1) \cdot \alpha_1 
	$.
\end{exa}
Adding negative constants is a mild extension to allow subtraction in the logic. 
It follows from our characterization of $\shp$ that this is exactly what we need to capture  $\gp$.
\begin{cor} \label{prop:capture-gapp}
	$\eqsoz(\fo)$ captures $\gp$ over ordered structures.
\end{cor}
This is an interesting result that shows how robust is $\qso$ when capturing different complexity classes.

A different class of functions comes from considering the optimization version of a decision problem. For example, one can define MAX-SAT as the problem of determining the maximum number of clauses, of a given CNF propositional formula, that can be made true by an assignment. Here, MAX-SAT is defined in terms of a maximization problem which in its essence differs from the functions in $\shp$. 
To formalize this set of optimization problems, Krentel defined $\optp$~\cite{krentel1988complexity} as the class of functions computable by taking the maximum or minimum of the output values over all runs of a polynomial-time NTM machine with output tape (i.e. each run produces a binary string which is interpreted as a number). 
For instance, MAX-SAT is in $\optp$ as many other optimization versions of $\np$-problems.
Given that in~\cite{krentel1988complexity} the author does not make the distinction between $\max$ and $\min$, in~\cite{vollmer1995complexity} the authors define the classes $\maxp$ and $\minp$ as the max and min version of the problems in $\optp$ (i.e. $\optp = \maxp \cup \minp$).

In order to capture classes of optimization functions, we extend as follows $\qso$ with $\max$ and $\min$ quantifiers (called $\optqso$). 
Given a signature $\R$, the set of $\optqso$-formulae over $\R$ is given by extending the syntax in (\ref{syntax}) with the following operators:
\begin{align*}
\max\{\alpha,\alpha\} \ \mid\ \min\{\alpha,\alpha\} \ \mid \maxa{x} \alpha \ \mid \ \mina{x} \alpha \ \mid \ \maxa{X} \alpha \ \mid \ \mina{X} \alpha 
\end{align*}
where $x \in \fv$ and $X \in \sv$. The semantics of the $\qso$-operators in $\optqso$ are defined as usual. Furthermore, the semantics of the max and min quantifiers are defined as the natural extension of the sum quantifiers in $\qso$ (see Table~\ref{tab-semantics}) by maximizing or minimizing, respectively, instead of computing a sum. 
\begin{exa}\label{ex:optqso}
	Recall again the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the size of the largest clique in a graph. This can be done in $\optqso$ as follows:
	\[
\alpha_6 := \maxa{X} \left( \, \clique(X) \cdot \sa{z} X(z)  \, \right)
	\]
	Notice that formula $\sa{z} X(z)$ is used to compute the number of nodes in a set $X$. 
\end{exa}
Similar than for $\maxp$ and $\minp$, we have to distinguished between the $\max$ and $\min$ fragments of $\optqso$. For this, we define the fragment $\maxqso$ of all $\optqso$ formulae constructed from $\qfo$ operators and $\max$-formulae $\max\{\alpha,\alpha\}$, $\maxa{x} \alpha$ and  $\maxa{X} \alpha$.
The class $\minqso$ is defined analogously changing $\max$ with $\min$. Notice that in $\maxqso$ and $\minqso$, second-order sum and product are not allowed. For instance, formula $\alpha_6$ in Example \ref{ex:optqso} is in $\maxqso$.
As one could expect, $\maxqso$ and $\minqso$ are the needed logics to capture $\maxp$ and $\minp$.
\begin{thm} \label{theo:capture-optp}
	$\maxqso(\fo)$ and $\minqso(\fo)$ capture $\maxp$ and $\minp$, respectively, over ordered structures.
\end{thm}
\proof
\input{proof-capture-optp}
\qed
It is important to mention that a similar result was proved in~\cite{kolaitis1994logical} for the class $\maxpb$ (resp., $\minpb$) of problems in $\maxp$ (resp., $\minp$) whose output value is polynomially bounded.
Interestingly, Theorem \ref{theo:capture-optp} is stronger since our logic has the freedom to use sum and product quantifiers, instead of using a max-and-count problem over Boolean formulae. 
Finally, it is easy to prove that our framework can also capture $\maxpb$ and $\minpb$ by disallowing the product $\Pi x$ in $\maxqso(\fo)$ and $\minqso(\fo)$, respectively.

>>>>

There exist complexity classes that do not fit in our framework because either the output of a function in not a natural number (e.g. a negative number) or the class is not defined purely in terms of arithmetical operations (e.g. min and max).
To remedy this problem, we show here how $\qso$ can be easily extended  to capture such classes that go beyond sum and product over natural numbers.

It is well-known that, under some reasonable complexity-theoretical assumptions, $\shp$ is not closed under subtraction, not even under subtraction by one~\cite{OH93}.
To overcome this limitation, $\gp$ was introduced in~\cite{FFK94} as the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x) - \tmr_M(x)$, where  $\tmr_M(x)$ is the number of rejecting runs of $M$ with input $x$.
That is, $\gp$ is the closure of $\shp$ functions under subtraction, and its functions can obviously take negative values.
Given that our logical framework was built on top of the natural numbers, we need to extend $\qso$ in order to capture $\gp$. 
The most elegant way to do this is by allowing constants coming from $\bbZ$ instead of just $\bbN$. 
Formally, we define the logic $\qsoz$ whose syntax is the same as in \eqref{syntax} and whose semantics is the same as in Table~\ref{tab-semantics} except that the atomic formula $s$ (i.e. a constant) comes from $\bbZ$.  
Similar than for $\qso$, we define the fragment $\eqsoz$ as the extension of $\eqso$ with constants in $\bbZ$.
\begin{exa}
	Recall the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the number of cliques in a graph that are not triangles. This can be easily done in $\qsoz$ with the formula:
	$
	\alpha_5 :=	\alpha_2 + (-1) \cdot \alpha_1 
	$.
\end{exa}
Adding negative constants is a mild extension to allow subtraction in the logic. 
It follows from our characterization of $\shp$ that this is exactly what we need to capture  $\gp$.
\begin{cor} \label{prop:capture-gapp}
	$\eqsoz(\fo)$ captures $\gp$ over ordered structures.
\end{cor}
This is an interesting result that shows how robust and versatile is $\qso$ for capturing different counting complexity classes even beyond $\bbN$.

A different class of functions comes from considering the optimization version of a decision problem. For example, one can define MAX-SAT as the problem of determining the maximum number of clauses, of a given CNF propositional formula, that can be made true by an assignment. Here, MAX-SAT is defined in terms of a maximization problem which in its essence differs from the functions in $\shp$. 
To formalize this set of optimization problems, Krentel defined $\optp$~\cite{krentel1988complexity} as the class of functions computable by taking the maximum or minimum of the output values over all runs of a polynomial-time NTM machine with output tape (i.e. each run produces a binary string which is interpreted as a number). 
For instance, MAX-SAT is in $\optp$ as many other optimization versions of $\np$-problems.
Given that in~\cite{krentel1988complexity} Krentel did not make the distinction between $\max$ and $\min$, in~\cite{vollmer1995complexity} they defined the classes $\maxp$ and $\minp$ as the max and min version of the problems in $\optp$ (i.e. $\optp = \maxp \cup \minp$).

In order to capture classes of optimization functions, we extend as follows $\qso$ with $\max$ and $\min$ quantifiers (called $\optqso$). 
Given a signature $\R$, the set of $\optqso$-formulae over $\R$ is given by extending the syntax in (\ref{syntax}) with the following operators:
\begin{align*}
\max\{\alpha,\alpha\} \ \mid\ \min\{\alpha,\alpha\} \ \mid \maxa{x} \alpha \ \mid \ \mina{x} \alpha \ \mid \ \maxa{X} \alpha \ \mid \ \mina{X} \alpha 
\end{align*}
where $x \in \fv$ and $X \in \sv$. The semantics of the $\qso$-operators in $\optqso$ are defined as usual. Furthermore, the semantics of the max and min quantifiers are defined as the natural extension of the sum quantifiers in $\qso$ (see Table~\ref{tab-semantics}) by maximizing or minimizing, respectively, instead of computing a sum or a product. 
\begin{exa}\label{ex:optqso}
	Recall again the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the size of the largest clique in a graph. This can be done in $\optqso$ as follows:
	\[
\alpha_6 := \maxa{X} \left( \, \clique(X) \cdot \sa{z} X(z)  \, \right)
	\]
	Notice that formula $\sa{z} X(z)$ is used to compute the number of nodes in a set $X$. 
\end{exa}
Similar than for $\maxp$ and $\minp$, we have to distinguished between the $\max$ and $\min$ fragments of $\optqso$. For this, we define the fragment $\maxqso$ of all $\optqso$ formulae constructed from $\qfo$ operators and $\max$-formulae $\max\{\alpha,\alpha\}$, $\maxa{x} \alpha$ and  $\maxa{X} \alpha$.
The class $\minqso$ is defined analogously changing $\max$ with $\min$. Notice that in $\maxqso$ and $\minqso$, second-order sum and product are not allowed. For instance, formula $\alpha_6$ in Example \ref{ex:optqso} is in $\maxqso$.
As one could expect, $\maxqso$ and $\minqso$ are the needed logics to capture $\maxp$ and $\minp$.
\begin{thm} \label{theo:capture-optp}
	$\maxqso(\fo)$ and $\minqso(\fo)$ capture $\maxp$ and $\minp$, respectively, over ordered structures.
\end{thm}
\proof
\input{proof-capture-optp}
\qed
It is important to mention that a similar result, following the framework of~\cite{SalujaST95}, was proved in~\cite{kolaitis1994logical} for the class $\maxpb$ (resp., $\minpb$) of problems in $\maxp$ (resp., $\minp$) whose output value is polynomially bounded.
Interestingly, Theorem \ref{theo:capture-optp} is stronger since our logic has the freedom to use sum and product quantifiers, instead of using a max-and-count problem over Boolean formulae. 
Finally, it is easy to prove that our framework can also capture $\maxpb$ and $\minpb$ by disallowing the product $\Pi x$ in $\maxqso(\fo)$ and $\minqso(\fo)$, respectively.

### 2017-07-24 LMCS/proof-capture-fp.tex
Paso completo a sección 4.

In the next proof and the followings, we will use the symbol $<$ also to denote the lexicographic order over same-sized tuples. Formally, for $\bar{x} = (x_1,\ldots,x_m)$ and $\bar{y} = (y_1,\ldots,y_m)$ we denote by $\bar{x} < \bar{y}$ the formula:
$$
\bigvee_{i = 1}^m \, \bigwedge_{j = 1}^{i-1} \, (x_j = y_j \wedge x_i < y_i).
$$
Similarly, we use $\bar{x} = \bar{y}$ to denote equality between tuples and $\bar{x} \leq \bar{y}$  to denote $\bar{x} < \bar{y} \vee \bar{x} = \bar{y}$.
%Furthermore, If $\bar{x} = (x_1,\ldots,x_m)$ ($\bar{X} = (X_1,\ldots,X_m)$) is a tuple of first-order (second-order) variables, we denote $\sa{\bar{x}}\alpha$ for $\sa{x_1}\cdots\sa{x_m}\alpha$ and $\sa{\bar{X}}\alpha$ for $\sa{X_1}\cdots\sa{X_m}\alpha$ for each $\qso$ formula $\alpha$.
We also use some syntactic sugar in $\qso$ to simplify formulas like the {\em conditional count} symbol $(\varphi \mapsto \alpha)$ defined as $ (\varphi\cdot\alpha) + \neg\varphi$ for any Boolean formula $\varphi$ and any quantitative formula $\alpha$. Note that for each $\A \in \ostr[\R]$, and each first-order (second-order) assignment $v$ ($V$) over $\A$:
\[
\sem{(\varphi \mapsto \alpha)}(\A,v,V) = 
\begin{cases}
\sem{\alpha}(\A,v,V) &\text{if } (\A,v,V)\models\varphi,\\
1 &\text{otherwise}.
\end{cases}
\]

Now we prove Theorem~\ref{theo:capture-fp}. For condition (1), recall that checking whether $\A\models\varphi$ for any $\lfp$ formula $\varphi$ can be done in deterministic polynomial time on the size of $\A$\cite{I83}. 
%Also, the constant function $s$ can be trivially simulated in $\fp$. 
Furthermore, it is easy to check that $\fp$ is closed under polynomial sum and multiplication. 
We conclude that any formula in $\qfo(\lfp)$ can be computed in $\fp$.	
For condition (2), let $\R$ be a signature, $f\in \fp$
and $\ell\in\nat$ such that $\log_2\left(f(\enc(\A)) \right) \leq \size{\A}^\ell$ for every $\A\in\ostr[\R]$ (i.e. $\size{\A}^\ell$ is an upper bound for the output size of $f$ over $\A$).
Consider the language:
\[
L = \{(\A,\bar{a})\mid \bar{a} \in A^l \text{ and the } \bar{a}\text{-th bit of }f(\enc(\A))\text{ is 1}\}.
\]
where $\bar{a}$ encodes a number by following the lexicographic order over $A^l$.
%Consider a procedure that receives $\enc(\A)$ and an assignment $\bar{a}$ to $\bar{x}$. Let $m$ be the position of $\bar{a}$ in the lexicographic order of the tuples in $A^{\ell}$. The procedure then computes the $m$-th bit of $f(\enc(\A))$, from least to most significant. 
Since this language is in $\ptime$, by \cite{I83} there exists a formula $\Phi(\bar{x})$ in $\lfp$ such that $\A\models\Phi(\bar{a})$ if and only if $(\A,\bar{a})\in L$. 
Then we use the following formula to encode $f$:
$$
\alpha = \sa{\bar{x}} \Phi(\bar{x})\cdot \pa{\bar{y}}(\bar{y} < \bar{x} \mapsto 2)
$$
Here, the subformula $\pa{\bar{y}}(\bar{y} < \bar{x} \mapsto 2)$ takes the value $2^m$ if there exist $m$ tuples in $A^{\ell}$ that are smaller than $\bar{x}$. Summing these values for each $\bar{a}\in A^{\ell}$ gives exactly $f(\enc(\A))$. 
In other words, $\Phi(\bar{x})$ simulates the behavior of the $\fp$-machine and the formula $\alpha$ reconstructs the binary output bit by bit.
Then, $\alpha$ is in $\qfo(\lfp)$ and $\sem{\alpha}(\A) = f(\enc(\A))$.

>>>>

In this proof and the followings, we will reuse the symbol $<$ to denote the lexicographic order over same-sized tuples. Formally, for $\bar{x} = (x_1,\ldots,x_m)$ and $\bar{y} = (y_1,\ldots,y_m)$ we denote by $\bar{x} < \bar{y}$ the formula:
$$
\bigvee_{i = 1}^m \, \bigwedge_{j = 1}^{i-1} \, (x_j = y_j \wedge x_i < y_i).
$$
Similarly, we use $\bar{x} = \bar{y}$ to denote equality between tuples and $\bar{x} \leq \bar{y}$  to denote $\bar{x} < \bar{y} \vee \bar{x} = \bar{y}$.
%Furthermore, If $\bar{x} = (x_1,\ldots,x_m)$ ($\bar{X} = (X_1,\ldots,X_m)$) is a tuple of first-order (second-order) variables, we denote $\sa{\bar{x}}\alpha$ for $\sa{x_1}\cdots\sa{x_m}\alpha$ and $\sa{\bar{X}}\alpha$ for $\sa{X_1}\cdots\sa{X_m}\alpha$ for each $\qso$ formula $\alpha$.
We will also use some syntactic sugar in $\qso$ to simplify formulas. Specifically, we will use the {\em conditional count} symbol $(\varphi \mapsto \alpha)$ defined as $ (\varphi\cdot\alpha) + \neg\varphi$ for any Boolean formula $\varphi$ and any quantitative formula $\alpha$. Note that for each $\A \in \ostr[\R]$, and each first-order (second-order) assignment $v$ ($V$) over $\A$:
\[
\sem{(\varphi \mapsto \alpha)}(\A,v,V) = 
\begin{cases}
\sem{\alpha}(\A,v,V) &\text{if } (\A,v,V)\models\varphi,\\
1 &\text{otherwise}.
\end{cases}
\]

Now we prove Theorem~\ref{theo:capture-fp}. For condition (1), recall that checking whether $\A\models\varphi$ for any $\lfp$ formula $\varphi$ can be done in deterministic polynomial time on the size of $\A$\cite{I83}. 
%Also, the constant function $s$ can be trivially simulated in $\fp$. 
Furthermore, it is easy to check that $\fp$ is closed under polynomial sum and multiplication. 
We conclude then that any formula in $\qfo(\lfp)$ can be computed in $\fp$.	
For condition~(2), let $\R$ be a signature, $f\in \fp$
and $\ell\in\nat$ such that $\log_2\left(f(\enc(\A)) \right) \leq \size{\A}^\ell$ for every $\A\in\ostr[\R]$ (i.e. $\size{\A}^\ell$ is an upper bound for the output size of $f$ over $\A$).
Consider the language:
\[
L = \{(\A,\bar{a})\mid \bar{a} \in A^l \text{ and the } \bar{a}\text{-th bit of }f(\enc(\A))\text{ is 1}\}.
\]
where $\bar{a}$ encodes a number by following the lexicographic order over $A^l$.
%Consider a procedure that receives $\enc(\A)$ and an assignment $\bar{a}$ to $\bar{x}$. Let $m$ be the position of $\bar{a}$ in the lexicographic order of the tuples in $A^{\ell}$. The procedure then computes the $m$-th bit of $f(\enc(\A))$, from least to most significant. 
Clearly, the language $L$ is in $\ptime$ and by \cite{I83} there exists a formula $\Phi(\bar{x})$ in $\lfp$ such that $\A\models\Phi(\bar{a})$ if, and only if, $(\A,\bar{a})\in L$. 
We use then the following formula to encode $f$:
$$
\alpha = \sa{\bar{x}} \Phi(\bar{x})\cdot \pa{\bar{y}}(\bar{y} < \bar{x}) \mapsto 2)$$
Note that the subformula $\pa{\bar{y}}(\bar{y} < \bar{x}) \mapsto 2$ takes the value $2^m$ if there exist $m$ tuples in $A^{\ell}$ that are smaller than $\bar{x}$. Summing these values for each $\bar{a}\in A^{\ell}$ gives exactly $f(\enc(\A))$. 
In other words, $\Phi(\bar{x})$ simulates the behavior of the $\fp$-machine and the formula $\alpha$ reconstructs the binary output bit by bit.
Then $\alpha$ is in $\qfo(\lfp)$ and $\sem{\alpha}(\A) = f(\enc(\A))$.

### 2017-07-24 LMCS/proof-capture-optp.tex
Paso completo a sección 4.

In this proof we construct recursively a non-deterministic polynomial time Turing machine $M_{\alpha}$ with output tape for each $\maxqso(\fo)$ formula $\alpha$ over a signature $\R$.
This machine, on input $(\A,v,V)$, non-deterministically produces $\sem{\alpha}(\A,v,V)$ (in binary) over the output tape of some run, and this value is the maximum value over all runs.
Suppose $\A$ has domain $A$. If $\alpha$ is a $\fo$-formula $\varphi$, then the machine checks if $(\A,v,V)\models\varphi$ in deterministic polynomial time, and output $1$ if, and only if, $(\A,v,V)$ satisfies $\varphi$. If $\alpha$ is a constant $s$, it outputs $s$ in binary over the output tape. If $\alpha = (\beta \add \gamma)$, then it simulates $M_{\beta}$ and $M_{\gamma}$ on input $(\A,v,V)$, adds the output of both machines and prints this value over the output tape. If $\alpha = (\beta \mult \gamma)$ or $\alpha = \max\{\beta,\gamma\}$, then it does the same than the previous case but it multiplies or maximizes, respectively, the outputs of both machines instead of adding. 
If $\alpha = \sa{x}\beta$, it iterates over all elements $a\in A$ simulating and adding the output of $M_{\beta}$ on input $(\A,v[a/x],V)$. $M_{\alpha}$ finally outputs the aggregated value in the output tape. 
If $\alpha = \pa{x}\beta$ or $\alpha = \maxa{x}\beta$, then $M_{\alpha}$ does the same idea than the previous case with the difference that the output of $M_{\beta}$ on input $(\A,v[a/x],V)$ is multiplied or maximized, respectively.
If $\alpha = \maxa{X}\beta$, it chooses $B\in A^{arity(X)}$ and simulates $M_{\beta}$ on input $(\A,v,V[B/X])$.
This covers all possible cases for $\alpha$.
Furthermore, it is straightforward to prove that each of these steps are correct and can be computed with a non-deterministic polynomial time Turing machine with output tape.
Let $\alpha$ be a formula in $\maxqso$ over a signature $\R$ and let $f$ be a function over $\R$ such that $f(\enc(\A))$ is equal to the maximum run (with respect to the output value) of $M_{\alpha}$ on input $(\A,v,V)$ for some $(\A,v,V) \in \ostr[\R]^*$. We have that $f$ is a $\maxp$ function over $\R$ and that $f(\enc(\A)) = \sem{\alpha}(\A)$ for every $\A\in\ostr[\R]$.
Finally, one can easily see that the same construction holds with $\minqso(\fo)$ by constructing a Turing machine that take the min over all runs instead of max.

The proof for the other direction is similar than in \cite{kolaitis1994logical} extended with the ideas of Theorem~\ref{theo:capture-fp}. Let $f\in \maxp$ be a function defined over some signature $\R$ and
$\ell\in\nat$ such that $\lceil\log_2 f(\enc(\A)) \rceil \leq n^\ell$ for each $\A\in\ostr[\R]$ of size $n$.
For $U \subseteq A^{\ell}$, we can interpret the encoding of $U$ ($\enc(U)$) as the binary encoding of a number with $n^l$-bits. We denote this value by $\val(\enc(U))$.
Then, given $\A\in\ostr[\R]$ and $U \subseteq A^{\ell}$, consider the problem of checking whether $f(\enc(\A)) \geq \val(\enc(U))$. 
Clearly, this is an $\np$-problem and, by Fagin's theorem, there exists a formula of the form $\ex{\bar{X}} \Phi(\bar{X}, Y)$ with $\Phi(\bar{X}, Y)$ in $\fo$-logic and $\arity(Y) = \ell$ such that $f(\enc(\A)) \geq \val(\enc(U))$ if, and only if, $(\A,v,V) \models \ex{\bar{X}} \Phi(\bar{X}, Y)$ with $V(Y) = U$. 
Then we can describe the function $f$ by the following $\maxqso$ formula:
$$
\alpha = \maxa{\bar{X}} \maxa{Y} \ \Phi(\bar{X}, Y) \cdot \big( \sa{\bar{x}} Y(\bar{x}) \cdot \pa{\bar{y}}(\bar{x} < \bar{y} \mapsto 2) \big).
$$
Note that, in contrast with previous proofs, we use $\bar{x} < \bar{y}$ instead of $\bar{y} < \bar{x}$ because the most significant bit in $\enc(U)$ correspond to the smallest tuple in $U$.  
It is easy to check that $\Phi(\bar{X}, Y)$ simulates the NP-machine and, if $\Phi(\bar{X}, Y)$ holds, the formula to the right  reconstructs the binary output from the relation in $Y$.
Then, $\alpha$ is in $\maxqso(\fo)$ over $\R$ and $\sem{\alpha}(\A) = f(\enc(\A))$.

For the case of $\minqso(\fo)$ and a function $f \in \minp$, one has to follow the same approach but considering the $\np$-problem of checking whether $f(\enc(\A)) \leq \val(\enc(U))$. Then, the formula for describing $f$ is the following:
$$
\alpha = \mina{\bar{X}} \mina{Y} \ \sa{\bar{x}} \big( (\Phi(\bar{X}, Y) \rightarrow Y(\bar{x})) \cdot \pa{\bar{y}}(\bar{x} < \bar{y} \mapsto 2)  \big).
$$
In this case, if the formula $\Phi(\bar{X}, Y)$ is false, then the output produced by the subformula inside the $\min$-quantifiers will be the biggest possible value (i.e. $2^{n^\ell}$).
On the other hand, if $\Phi(\bar{X}, Y)$ holds, the subformula will produce $\val(\enc(U))$. 
Similar than for $\max$, we conclude that $\alpha$ is in $\minqso(\fo)$ and $\sem{\alpha}(\A) = f(\enc(\A))$.

>>>>

It is straightforward to prove that $\maxp$ can compute any $\fo$-formula, is closed under first-order sum and product, and second-order maximization. 
Therefore, condition (1) in Definition~\ref{def:cap} follows similar than in the previous characterizations. Furthermore, one can easily see that the same holds with $\minqso(\fo)$.
The proof for the other direction is similar than in \cite{kolaitis1994logical} extended with the ideas of Theorem~\ref{theo:capture-fp}. Let $f\in \maxp$ be a function defined over some signature $\R$ and
$\ell\in\nat$ such that $\lceil\log_2 f(\enc(\A)) \rceil \leq |\A|^\ell$ for each $\A\in\ostr[\R]$.
For $U \subseteq A^{\ell}$, we can interpret the encoding of $U$ ($\enc(U)$) as the binary encoding of a number with $|\A|^\ell$-bits. We denote this value by $\val(\enc(U))$.
Then, given $\A\in\ostr[\R]$ and $U \subseteq A^{\ell}$, consider the problem of checking whether $f(\enc(\A)) \geq \val(\enc(U))$. 
Clearly, this is an $\np$-problem and, by Fagin's theorem, there exists a formula of the form $\ex{\bar{X}} \Phi(\bar{X}, Y)$ with $\Phi(\bar{X}, Y)$ in $\fo$ and $\arity(Y) = \ell$ such that $f(\enc(\A)) \geq \val(\enc(U))$ if, and only if, $(\A,v,V) \models \ex{\bar{X}} \Phi(\bar{X}, Y)$ with $V(Y) = U$. 
Then we can describe $f$ by the following $\maxqso$ formula:
$$
\alpha = \maxa{\bar{X}} \maxa{Y} \ \Phi(\bar{X}, Y) \cdot \big( \sa{\bar{x}} Y(\bar{x}) \cdot \pa{\bar{y}}(\bar{x} < \bar{y} \mapsto 2) \big).
$$
Note that, in contrast with previous proofs, we use $\bar{x} < \bar{y}$ instead of $\bar{y} < \bar{x}$ because the most significant bit in $\enc(U)$ correspond to the smallest tuple in $U$.  
It is easy to check that $\Phi(\bar{X}, Y)$ simulates the NP-machine and, if $\Phi(\bar{X}, Y)$ holds, the formula to the right  reconstructs the binary output from the relation in $Y$.
Then, $\alpha$ is in $\maxqso(\fo)$ over $\R$ and $\sem{\alpha}(\A) = f(\enc(\A))$.

For the case of $\minqso(\fo)$ and a function $f \in \minp$, one has to follow the same approach but considering the $\np$-problem of checking whether $f(\enc(\A)) \leq \val(\enc(U))$. Then, the formula for describing $f$ is the following:
$$
\alpha = \mina{\bar{X}} \mina{Y} \ \sa{\bar{x}} \big( (\Phi(\bar{X}, Y) \rightarrow Y(\bar{x})) \cdot \pa{\bar{y}}(\bar{x} < \bar{y} \mapsto 2)  \big).
$$
In this case, if the formula $\Phi(\bar{X}, Y)$ is false, then the output produced by the subformula inside the $\min$-quantifiers will be the biggest possible value (i.e. $2^{{|\A|}^\ell}$).
On the other hand, if $\Phi(\bar{X}, Y)$ holds, the subformula will produce $\val(\enc(U))$. 
Similar than for $\max$, we conclude that $\alpha$ is in $\minqso(\fo)$ and $\sem{\alpha}(\A) = f(\enc(\A))$.
